% !TEX program = pdflatex
% Draft v0.1 — venue TBD; using generic article layout, swap class later.
\documentclass[11pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{xcolor}
\usepackage[colorlinks=true,linkcolor=blue,citecolor=blue,urlcolor=blue]{hyperref}
\usepackage{CJKutf8} % for Chinese scenario names in examples; swap to xeCJK if compiling with xelatex

% ---- system name placeholder: final name TBD ----
\newcommand{\sysname}{\textsc{MemoirWorld}}
\newcommand{\todo}[1]{\textcolor{red}{[TODO: #1]}}

\title{\sysname: Consensus-Compressed Shared Memory for\\ Scalable Story-World Agent Societies}
\author{Yu Pan\thanks{\todo{author list, affiliation, email}}}
\date{}

\begin{document}
\maketitle

\begin{abstract}
Large language model (LLM) agent societies that role-play long-running story worlds
must remember what happened --- yet the dominant designs either store every
observation per agent, duplicating shared experiences across the cast, or
periodically summarize histories in ways that erase the very details that make
characters believable. We present \sysname, a multi-agent framework whose
long-term memory is a single shared vector store in which each atomic memory
entry carries an explicit \emph{owner set}. A normalization gate splits or
compresses every incoming memory into atomic, length-bounded statements, and a
\emph{consensus} mechanism merges semantically equivalent entries across agents,
keeping the shorter surface form, unioning owners, and preserving the earlier
story-time stamp. Combined with an event-driven tick scheduler in which idle
agents cost nothing, \sysname{} supports societies of dozens of characters,
including \emph{archived} agents for deceased characters whose memories remain
part of the shared record. A two-pass \emph{history sedimentation} pipeline
initializes a society directly from the first part of a novel (or a real-world
timeline): a registry pass canonicalizes characters, aliases, and places under
closed-world constraints, and a per-chapter, per-character exhaustive extraction
pass deposits time-stamped memories through the same consensus channel, so that
shared experiences (e.g., an oath sworn by three characters) converge to single
multi-owner entries. Agents then bootstrap their own goals from recalled
memories and simulate the \emph{sequel}, which we evaluate against the actual
continuation of the source. Across four scenarios --- two classical Chinese
novels, one English classic, and a real-world geopolitical timeline --- we
compare against Generative Agents, G-Memory, and Collaborative Memory under
identical brains, budgets, and schedulers, measuring memory footprint,
wall-clock and
token cost, narrative quality, continuation fidelity against ground truth, and
memory-grounded question answering. \sysname{} reaches comparable or better
quality at a fraction of the memory footprint and cost, and its memory dynamics
are directly interpretable: additions, merges, and revisions align with plot
events and character goals. \todo{fill in headline numbers}
\end{abstract}

\section{Introduction}
\label{sec:intro}

LLM-driven agent societies can now sustain multi-character role-play over long
horizons: characters converse, pursue goals, move through a world, and
accumulate experience. The bottleneck is memory. A society of $N$ characters
that each privately logs every turn stores the same shared scene $N$ times;
retrieval slows, context windows fill with near-duplicates, and cost grows with
the square of social activity. Conversely, aggressive summarization keeps
memory small but destroys the specific, character-bound details --- who swore
what to whom, and when --- that story worlds live on.

We argue that the unit of memory in a \emph{society} should not be the private
log but the \emph{shared fact}. Most of what characters remember, they remember
together: the oath in the peach garden is one fact with three owners, not three
private paraphrases. \sysname{} builds this observation into the storage layer
itself.

\paragraph{Contributions.}
\begin{enumerate}
  \item \textbf{Consensus-compressed shared memory.} A single vector store of
  atomic, length-bounded entries with explicit owner sets; inserts pass a
  normalization gate (split/compress into atoms) and a consensus check
  (embedding pre-filter + LLM equivalence) that merges duplicates across
  agents, keeps the shorter text, unions owners, and retains the earlier
  story-time. Deletion and revision are owner-set operations, so one agent
  forgetting does not erase a fact others still hold. (\S\ref{sec:memory})
  \item \textbf{History sedimentation.} A two-pass pipeline that turns the
  first part of a book or a real-world timeline into a society: a closed-world
  registry of canonical characters/places/aliases, then per-chapter,
  per-character exhaustive extraction of time-stamped atomic memories deposited
  through the consensus channel. Dead characters become \emph{archived} agents:
  never scheduled, but their memories persist in the shared record.
  Characters start the sequel with \emph{empty goal stacks} and bootstrap
  their own goals from recall --- goals are outcomes, not inputs.
  (\S\ref{sec:sedimentation})
  \item \textbf{An evaluation protocol for sequel simulation.} Beyond
  LLM-judged narrative quality, we score \emph{continuation fidelity} against
  the actual later part of the source (event hit rate, trajectory consistency)
  and \emph{memory-grounded QA} accuracy, and we report full memory/cost
  curves, enabling quality-per-byte comparisons. (\S\ref{sec:setup})
  \item \textbf{Evidence.} Across five diverse scenarios and three strong
  memory baselines under a strictly controlled runtime, \sysname{} attains
  similar or better quality with \todo{X$\times$} smaller memory and
  \todo{Y$\times$} lower cost, with interpretable memory dynamics.
  (\S\ref{sec:results}--\S\ref{sec:case})
\end{enumerate}

\section{Related Work}
\label{sec:related}

\paragraph{Memory for LLM agents.}
Generative Agents~\cite{park2023generative} introduced the memory stream:
append-only observations with recency/importance/relevance retrieval and
periodic reflection. MemGPT~\cite{packer2023memgpt} pages memory between
context and external storage. These designs are per-agent: in a society,
shared experiences are duplicated per participant.

\paragraph{Multi-agent memory systems.}
G-Memory~\cite{gmemory2025} maintains hierarchical insight/query/interaction
graphs shared across agents, inspired by organizational memory theory.
Collaborative Memory~\cite{collabmem2025} shares memory \emph{fragments}
across users and agents under dynamic bipartite access-control policies with
immutable provenance --- the design closest to ours, but it governs
\emph{who may read} a fragment, whereas we govern \emph{how many copies
exist}: consensus dedup, atomicity, and story-time provenance are absent.
LEGOMem~\cite{legomem2025} distributes modular \emph{procedural} memory
(task trajectories) between orchestrators and workers for workflow
automation; DAMCS~\cite{damcs2025} couples a decentralized adaptive
knowledge-graph memory with structured communication for embodied
cooperation. Both target task execution rather than long-horizon social
memory. Our approach differs in \emph{where} deduplication happens: not as a
posterior summarization step over private logs, but at insert time in the
storage layer, preserving atomic facts with provenance (owner sets, story
time) instead of prose summaries.

\paragraph{Story worlds and role-play societies.}
BookWorld~\cite{chen2025bookworld} constructs role-agent societies from
novels with a world agent that distills consensus summaries while compressing
per-role histories. BOOKMARKS~\cite{bookmarks2026} maintains per-role
storyline ``bookmarks'' for long-form role-play consistency, and
SALM~\cite{salm2025} reports sublinear memory growth in social-network
simulation platforms; both keep memory per-role rather than as a shared,
owner-attributed store. \todo{optional: add FURINA (role-play benchmark) and
a survey citation if space permits}

\paragraph{Simulation grounded in real timelines.}
\todo{brief para: social simulacra / world-model sims / wargaming with LLMs;
position the Russia--Ukraine scenario as continuation-fidelity benchmark
rather than forecasting claim.}

\section{The \sysname{} Framework}
\label{sec:framework}

\subsection{Runtime: agents, actions, and the tick barrier}
\label{sec:runtime}
An agent is one of three kinds --- \emph{character}, \emph{environment}, or
\emph{information carrier} --- with a pluggable brain (LLM, rules, or
retrieval). Each character holds a four-part short-term memory: a FIFO cache of
the last $K{=}20$ action--result pairs, a goal \emph{stack} (deeper goals more
fundamental), a status register with public/private keys, and a message inbox.
A global clock advances in ticks; in each tick every \emph{awake} agent decides
exactly one action (decisions run concurrently; effects apply in deterministic
agent order), messages sent at $t$ are delivered at $t{+}1$, and an agent whose
inbox and goal stack are both empty sleeps at zero cost until messaged.
Movement between environment nodes takes map-distance ticks. The action space
(21 actions) spans messaging (\texttt{say}, \texttt{gesture}), perception
(\texttt{observe}, \texttt{read}), cognition (\texttt{think},
\texttt{conclude}), goal and status management, memory operations
(\texttt{remember}, \texttt{recall}, \texttt{forget},
\texttt{revise\_memory}), and movement; agents are taught usage and standard
pipelines through a skill document injected into the system prompt.
\todo{figure: architecture diagram}

\subsection{Consensus-compressed shared long-term memory}
\label{sec:memory}
All long-term memory lives in one vector collection. An entry is a tuple
$(x, O, \tau, s)$: atomic text $x$ with $|x|\le L$ ($L{=}80$ chars by default),
owner set $O$, story-order $\tau$, and provenance $s$.

\paragraph{Normalization gate.} \texttt{remember}$(a, x)$ first checks $x$ for
length and atomicity (multiple sentence terminators or coordinating
connectives); violating inputs are split/compressed by one LLM call into atoms
of at most $L$ characters. Compliant inputs pass with zero LLM cost.

\paragraph{Consensus insert.} For each atom, retrieve the top-$k$ neighbors; if
none exceeds cosine similarity $\theta{=}0.86$, insert a new entry with
$O{=}\{a\}$. Otherwise a single LLM call judges semantic equivalence against
the candidates; on a match the store keeps the \emph{shorter} of the two
texts, sets $O \leftarrow O \cup \{a\}$, and keeps the \emph{earlier}
story-order (and its story-time label). Thus repeated tellings converge to one
canonical atom with provenance, and the shared-memory ratio
$|\{e : |O_e|\ge 2\}| / |E|$ becomes a measurable structural property of the
society.

\paragraph{Deletion and revision as owner-set operations.}
\texttt{forget} removes the caller from $O$ and deletes the entry only when
$O=\emptyset$; \texttt{revise\_memory} is forget-then-reinsert through the same
gate and consensus. \texttt{recall} is owner-filtered similarity search: agents
can only retrieve what they own, so information asymmetry between characters is
a storage-level invariant rather than a prompt-level convention.

\subsection{History sedimentation: initializing a society from a source text}
\label{sec:sedimentation}
Given the first part of a source (chapters $1..m$ of a novel; the first third
of a timeline), \sysname{} builds the society in two passes.
\textbf{Pass 1 (registry).} Chunk summaries feed one merge call that outputs a
closed-world registry: canonical IDs, display names, and \emph{aliases} for
characters, places, and information carriers. The registry is a reviewable
artifact; all later attribution is restricted to its IDs (unknown mentions are
either canonicalized through a resolver over names and aliases, added by a
single registry-completion call, or dropped with a warning --- never silently
invented).
\textbf{Pass 2 (sedimentation).} Chapter-aligned chunks are processed in story
order. For each chunk, a roster call lists the characters on stage and state
updates (location, alive); then \emph{one call per character} exhaustively
enumerates that character's experiences in the chunk --- actions, dialogue
gists, observations, relationship changes, gains and losses, stance shifts ---
as time-prefixed atoms; a coverage-audit call adds missed details. Every atom
is deposited via \texttt{remember} with story-order $\tau$, so shared scenes
merge into multi-owner entries at load time. Final states decide who is alive;
the dead become archived agents (never scheduled, not observable, memories
retained). Location handling is guarded by hard invariants (ASCII canonical
IDs, no duplicate names/aliases, every reference resolvable) enforced in code,
not prompts. The sediment is exported holographically (text + embeddings +
metadata), so societies rebuild with zero re-embedding.

\paragraph{Goal bootstrapping.} Sequel characters start with empty goal
stacks. On their first turns the view carries a bootstrap hint; agents recall
their past, observe their surroundings, conclude, and push their own
fundamental and immediate goals. In our runs, characters reconstructed
canonical motivations (e.g., \begin{CJK}{UTF8}{gbsn}曹操\end{CJK} deriving
``march on Xuzhou to avenge my father'' purely from sedimented memory and a
terse kickoff). \todo{tighten example after final runs}

\section{Experimental Setup}
\label{sec:setup}

\subsection{Scenarios}
\label{sec:scenarios}
Four societies spanning language, genre, cast size, and fiction vs.\ reality
(Table~\ref{tab:scenarios}). For each, the \emph{sediment} span initializes
memory; simulation then runs the sequel, and the held-out \emph{reference} span
provides ground truth for continuation fidelity. Sources are public-domain
texts (Wikisource) and CC-licensed timeline articles (Wikipedia), enabling
artifact release.

\begin{table}[t]
\centering\small
\begin{tabular}{lllll}
\toprule
Scenario & Language & Cast & Sediment & Reference \\
\midrule
Romance of the Three Kingdoms & zh & $\sim$40 & ch.~1--40 & ch.~41--60 \\
Dream of the Red Chamber & zh & $\sim$30 & ch.~1--40 & ch.~41--80 \\
War and Peace & en & $\sim$50 & Vol.~1--2 & Vol.~3 \\
Russia--Ukraine timeline & en & $\sim$25 & 2022/02--2023/06 & 2023/07--2024/12 \\
\bottomrule
\end{tabular}
\caption{Scenarios. Cast counts are registry sizes after curation; the
timeline scenario anthropomorphizes states, leaders, and organizations, with
key documents as information carriers. \todo{finalize counts}}
\label{tab:scenarios}
\end{table}

\subsection{Baselines}
\label{sec:baselines}
All baselines are re-implemented as \emph{memory backends} inside the \sysname{}
runtime: identical brains, action space, scheduler, kickoffs, and budgets ---
only the memory layer varies. The three chosen backends span the design space:
\textbf{Generative Agents}~\cite{park2023generative} (per-agent append-only
stream with recency/importance/relevance retrieval and periodic reflection ---
the canonical private-log design);
\textbf{G-Memory}~\cite{gmemory2025} (shared hierarchical
insight/query/interaction graph memory --- the recent structured-sharing
design);
\textbf{Collaborative Memory}~\cite{collabmem2025} (shared fragments with
provenance under per-agent access policies --- the closest access-controlled
sharing design, which controls visibility but does not deduplicate).
Sedimentation feeds each backend the same source span through its own
ingestion mechanism. \todo{document per-backend ingestion adaptations in
appendix; optionally add BOOKMARKS as a story-domain per-role baseline}

\subsection{Metrics}
\label{sec:metrics}
\textbf{Memory.} Entries and bytes over ticks (peak/final), per-agent vs.\
shared split, compression ratio against the raw event log.
\textbf{Time/cost.} Wall-clock per tick, retrieval latency, LLM calls and
tokens by purpose bucket; one-time sedimentation cost reported separately.
Quality has three groups (five metrics after merging redundant ones).
\textbf{(A) Narrative quality} (LLM-judge, 1--7): character fidelity and
coherence, following BookWorld's rubric for comparability; continuation
plausibility folds into coherence (both ask whether the produced text is
self-consistent with the preceding narrative). \todo{confirm exact BookWorld
dims and keep the overlapping subset}
\textbf{(B) Continuation fidelity} (against the held-out reference span, our
key novel axis): (i) \emph{Event Hit Rate} --- a checklist of key events mined
from the reference (e.g., ``Liu Bei relieves Tao Qian''), scored as the
fraction realized in simulation by a judge with human spot checks;
(ii) \emph{Character Trajectory Consistency} --- alignment of each character's
simulated goals/actions with their actual later arc.
\textbf{(C) Memory faithfulness}: \emph{Memory QA} --- $N{=}100$ factual
questions auto-generated from the sediment span, answered only via
\texttt{recall}; accuracy measures how well compressed memory preserves facts.
Judges use a different model family than the simulated brains; all prompts and
rubrics released. \todo{judge model choice; inter-rater stats}

\subsection{Protocol}
Each scenario $\times$ backend: identical kickoff and tick budget (50 ticks),
3 seeds, mean$\pm$std. Primary figures are Pareto fronts of quality vs.\ final
memory size and vs.\ cumulative token cost. \todo{seeds/ticks may increase for
camera-ready}

\section{Results}
\label{sec:results}
\todo{Main table: 4 scenarios $\times$ 4 systems $\times$ (memory, cost, quality
triplet). Pareto figures (quality vs.\ memory; quality vs.\ cost). Memory-vs-%
cast-size trend across the four scenarios. Sleep-economy analysis showing cost
vs.\ number of active agents.}

\section{Case Study: Interpretable Memory Dynamics}
\label{sec:case}
\todo{Three Kingdoms, 50-tick sequel: (a) timeline of one character's memory
adds/merges/revisions aligned against plot beats and the goal stack (Cao Cao:
revenge campaign); (b) a consensus merge example --- the peach-garden oath
converging to one three-owner entry; (c) sediment-density heatmap (character
$\times$ chapter) showing memory tracking narrative structure.}

\paragraph{Shared-memory structure vs.\ social behavior.}
We compute an agent$\times$agent \emph{shared-memory matrix} $M$, where
$M_{ii}$ is agent $i$'s entry count and $M_{ij}$ ($i\neq j$) the number of
entries jointly owned by $i$ and $j$. We correlate $M_{ij}$ against two
behavioral signals from the same run: communication volume (\texttt{say}/%
\texttt{gesture} counts between $i$ and $j$) and co-location time (ticks $i$
and $j$ spend in the same environment). The hypothesis --- that shared memory
grows with social contact --- is testable directly from the owner sets, event
log, and presence index, and quantifies the interpretability of consensus
merging. \todo{report Pearson/Spearman $r$ per scenario; heatmap figure of $M$;
scatter of $M_{ij}$ vs.\ comm and co-location.}

\paragraph{Information asymmetry.}
Owner-filtered \texttt{recall} makes stale-belief behavior an emergent property
(characters act on outdated knowledge until news arrives); we give one
concrete trace. \todo{pick the cleanest example from the 50-tick run.}

\section{Ablations}
\label{sec:ablations}
\todo{On Three Kingdoms (full grid) and consensus on/off everywhere:
consensus on/off; normalization gate on/off; $L \in \{40, 80, 160\}$;
recall top-$k \in \{3,5,10\}$; FIFO size $\{10,20,40\}$; goal bootstrap on/off;
archived agents kept vs.\ removed. Report memory/quality deltas.}

\section{Limitations and Ethics}
\label{sec:limitations}
Sedimentation quality depends on the extractor LLM; registry and life/death
states benefit from light human curation (we quantify curation effort).
The real-world scenario evaluates \emph{continuation fidelity on held-out past
events}; we make no forecasting claims, and we discuss the ethics of simulating
ongoing conflicts. \todo{expand}

\section{Conclusion}
\todo{write last}

\bibliographystyle{plain}
\begin{thebibliography}{9}
\bibitem{park2023generative} J.~S. Park et al. Generative Agents: Interactive
Simulacra of Human Behavior. UIST 2023.
\bibitem{packer2023memgpt} C.~Packer et al. MemGPT: Towards LLMs as Operating
Systems. arXiv:2310.08560.
\bibitem{chen2025bookworld} Y.~Chen et al. BookWorld: From Novels to
Interactive Agent Societies for Creative Story Generation. ACL 2025.
\todo{verify author list}
\bibitem{gmemory2025} G-Memory: A Hierarchical Graph Memory System for
Multi-Agent Systems Inspired by Organizational Memory Theory.
arXiv:2506.07398. \todo{authors}
\bibitem{collabmem2025} Collaborative Memory: Multi-User Memory Sharing in
LLM Agents with Dynamic Access Control. arXiv:2505.18279. \todo{authors}
\bibitem{legomem2025} LEGOMem: Modular Procedural Memory for Multi-agent LLM
Systems for Workflow Automation. arXiv:2510.04851. \todo{authors}
\bibitem{damcs2025} Decentralized Adaptive Knowledge Graph Memory and
Structured Communication for Multi-Agent Cooperation (DAMCS).
arXiv:2502.05453. \todo{authors}
\bibitem{salm2025} SALM: A Multi-Agent Framework for Language Model-Driven
Social Network Simulation. arXiv:2505.09081. \todo{authors}
\bibitem{bookmarks2026} BOOKMARKS: Efficient Active Storyline Memory for
Role-playing. arXiv:2605.14169. \todo{authors}
\end{thebibliography}

\end{document}
